---
title: "fastStructure"
author: "Harly Durbin"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 7)
library(readr)
library(dplyr)
library(stringr)
library(magrittr)
library(rlang)
library(purrr)
library(ggplot2)
library(tidylog)

source(here::here("source_functions/structure_functions.R"))
```

# Notes & questions

## Dataset key

* Which samples

```{r}
tibble::tribble(
  ~ Name, ~ Description, ~ n, ~ `k ran`,
  "full", "All species. All available samples.", "2,291","1-20",
  "20all", "All species. Up to 20 (ranked by coverage) samples per population label.", "886", "1-20",
  "5regionall", "All species. Up to 5 (ranked by coverage) samples per species per region", "142", "1-20",
  "20cattle", "Bos taurus, B.indicus, B. primigenius, composite. Up to 20 (ranked by coverage) samples per population label.", "768", "1-20",
  "5cattle", "Bos taurus, B.indicus, B. primigenius, composite. Up to 5 (ranked by coverage) samples per population label.", "392", "1-20",
  "allyak", "B. grunniens and B. mutus. All available samples.", "231", "1-10",
  "allgayalgaur", "B. frontalis and B. gaurus. All available samples.", "24", "1-10",
  "allbanteng", "B. javanicus. All available samples.", "8", "1-8"
) %>% 
  kableExtra::kable("html") %>% 
  kableExtra::kable_styling(full_width = F)
```

* Thinning parameter

```{r}
tibble::tribble(
  ~ Name, ~ Description,
  "superthin", "Randomly retain 0.001% of SNPs from each chromosome",
  "onepercent", "Randomly retain 0.01% of SNPs from each chromosome"
) %>% 
  kableExtra::kable("html") %>% 
  kableExtra::kable_styling(full_width = F)
```

# Setup

```{r}
sample_metadata <- 
  read_csv(here::here("data/derived_data/metadata/bovine_demo.sample_metadata.csv")) %>% 
  mutate(international_id = if_else(international_id == "CPC98_Bprimigenius_EnglandDerbyshire_5936", "ancient_derbyshire", international_id))
```

# Full datasets

* As expected, indicine/taurine ancestry in gayal and banteng ancestry in some East Asian indicine/composite + low levels of East Asian taurine <--> yak introgression

## `20all.onepercent` {.tabset}

* Model complexity that maximizes marginal likelihood = 4
* Model components used to explain structure in data = 6

### K = 4

```{r}

read_structure(dataset_thinning = "20all.onepercent", k = 4) %>%
  mutate(
    color_var =
      case_when(
        # Banteng, gaur, gayal
        variable == "X1" ~ "#fedb11",
        # Yaks
        variable == "X4" ~ "#c6b7d5",
        # European taurine
        variable == "X2" ~ "#126180",
        # Indian indicine
        variable == "X3" ~ "#ff7436"
      )
  ) %>%
  ggstructure(wrap_var = "species", custom_color = TRUE) +
  labs(title = "All samples, 20 samples per population, 1% SNPs: K = 4")

ggsave(here::here("figures/faststructure/20all.onepercent.K4.png"), width = 10, height = 7)
```

### K = 5

```{r}

read_structure(dataset_thinning = "20all.onepercent", k = 5) %>%
  mutate(
    color_var =
      case_when(
        # Banteng, gaur, gayal
        variable == "X1" ~ "#fedb11",
        # Yaks
        variable == "X2" ~ "#c6b7d5",
        # European taurine
        variable == "X3" ~ "#126180",
        # Indian indicine
        variable == "X4" ~ "#ff7436",
        # auroch
        variable == "X5" ~ "#84d6d3"
      )
  ) %>%
  ggstructure(wrap_var = "species", custom_color = TRUE) +
  labs(title = "All samples, 20 samples per population, 1% SNPs: K = 5")

ggsave(here::here("figures/faststructure/20all.onepercent.K5.png"), width = 10, height = 7)
```

### K = 6

```{r}
read_structure(dataset_thinning = "20all.onepercent", k = 6) %>%
  mutate(
    color_var =
      case_when(
        # European taurines
        variable == "X1" ~ "#126180",
        # Other taurines
        variable == "X3" ~ "#5f1f29",
        # Banteng, gaur, gayal
        variable == "X4" ~ "#fedb11",
        # Yaks
        variable == "X5" ~ "#c6b7d5",
        # Indian indicine
        variable == "X6" ~ "#ff7436",
      )
  ) %>%
  ggstructure(wrap_var = "species", custom_color = TRUE) +
  labs(title = "All samples, 20 samples per population, 1% SNPs: k = 6")



ggsave(here::here("figures/faststructure/20all.onepercent.K6.png"), width = 10, height = 7)
```

* The sixth component is probably a technical artifact, as it's less than zero (1e-06, which is why it isn't visible) for only one individual

```{r, echo=TRUE}
read_structure(dataset_thinning = "20all.onepercent", k = 6) %>% 
  filter(variable == "X2") %>% 
  filter(value != 0)
```

* Maybe interestingly, the Siberian auroch and non-European (particularly African & Asian steppe) taurines create their own component here (in maroon above)

```{r, echo=TRUE}
read_structure(dataset_thinning = "20all.onepercent", k = 6) %>% 
  filter(variable == "X3") %>% 
  filter(value != 0) %>% 
  arrange(desc(value), continent, region, population) %>% 
  select(variable, value, everything())
```

## `5regionall` {.tabset}

* Model complexity that maximizes marginal likelihood = 7
* Model components used to explain structure in data = 5


### K = 5

```{r}
read_structure(dataset_thinning = "5regionall.onepercent", k = 5) %>%
  mutate(
    color_var =
      case_when(
        # Indian indicine
        variable == "X1" ~ "#ff7436",
        # European taurine
        variable == "X2" ~ "#126180",
        # Gaur, gayal, bison
        variable == "X3" ~ "#fedb11",
        # Banteng
        variable == "X4" ~ "#8daed7",
        # Yaks
        variable == "X5" ~ "#c6b7d5",
      )
  ) %>%
  ggstructure(wrap_var = "species", custom_color = TRUE) +
  labs(title = "All samples, 5 samples per species per region, 1% SNPs: K = 5")

ggsave(
  here::here("figures/faststructure/5regionall.onepercent.K5.png"),
  width = 10,
  height = 7
)
```

### K = 6

```{r}
read_structure(dataset_thinning = "5regionall.onepercent", k = 6) %>%
  mutate(
    color_var =
      case_when(
        # Indian indicine
        variable == "X1" ~ "#ff7436",
        # Banteng, gaur, gayal
        variable == "X2" ~ "#8daed7",
        # Bison
        variable == "X3" ~ "#fedb11",
        # European taurine
        variable == "X5" ~ "#126180",
        # Yaks
        variable == "X6" ~ "#c6b7d5"
      )
  ) %>%
  ggstructure(wrap_var = "species", custom_color = TRUE) +
  labs(title = "All samples, 5 samples per species per region, 1% SNPs: K = 6")

ggsave(
  here::here("figures/faststructure/5regionall.onepercent.K6.png"),
  width = 10,
  height = 7
)
```

* Only 5 components actually used 

### K = 7

```{r}
read_structure(dataset_thinning = "5regionall.onepercent", k = 7) %>%
  mutate(
    color_var =
      case_when(
        # Banteng, gaur, gayal
        variable == "X1" ~ "#8daed7",
        # Indian indicine
        variable == "X2" ~ "#ff7436",
        # Yaks
        variable == "X3" ~ "#c6b7d5",
        # Bison
        variable == "X4" ~ "#fedb11",
        # European taurine
        variable == "X6" ~ "#126180",
      )
  ) %>%
  ggstructure(wrap_var = "species", custom_color = TRUE) +
  labs(title = "All samples, 5 samples per species per region, 1% SNPs: K = 7")

ggsave(
  here::here("figures/faststructure/5regionall.onepercent.K7.png"),
  width = 10,
  height = 7
)
```

* Again, only 5 components actually used 

# Cattle datasets 

* Find indicine ancestry in expected European breeds (Italian/Podolian breeds; Charolais) plus a few continental European individuals whose breeds have open herdbooks in the U.S
* Chinese indicines separate out at `20cattle.onepercent` K = 3,4,6 + `5cattle.onepercent` K = 4

## `20cattle.onepercent` {.tabset}

* Model complexity that maximizes marginal likelihood = 3
* Model components used to explain structure in data = 6


### K = 2

```{r}

read_structure(dataset_thinning = "20cattle.onepercent", k = 2) %>% 
  mutate(color_var =
           case_when(
             # taurine
             variable == "X1" ~ "#126180",
             # indicine
             variable == "X2" ~ "#ff7436"
           )) %>% 
  ggstructure(wrap_var = "region", custom_color = TRUE) +
  labs(title = "Cattle & ancient, 20 samples per population, 1% SNPs: k = 2")

ggsave(here::here("figures/faststructure/20cattle.onepercent.K2.png"), width = 10, height = 7)
```

* Find indicine ancestry in expected European breeds (Italian/Podolian breeds; Charolais) plus a few continental European individuals whose breeds have open herdbooks in the U.S.

```{r}
read_structure(dataset_thinning = "20cattle.onepercent", k = 2) %>% 
  filter(variable == "X2") %>% 
  filter(continent == "Europe") %>% 
  filter(value > 0.01) %>% 
  arrange(desc(value)) %>% 
  select(variable, value, everything())

```


### K = 3

```{r}

read_structure(dataset_thinning = "20cattle.onepercent", k = 3) %>%
  mutate(
    color_var =
      case_when(
        # European taurine
        variable == "X3" ~ "#126180",
        # Indian indicine
        variable == "X2" ~ "#ff7436",
        # Asian indicine
        variable == "X1" ~ "#629d62",
      )
  ) %>%
  ggstructure(wrap_var = "region", custom_color = TRUE) +
  labs(title = "Cattle & ancient, 20 samples per population, 1% SNPs: k = 3")

ggsave(here::here("figures/faststructure/20cattle.onepercent.K3.png"), width = 10, height = 7)

```

* New green component separates out East Asian indicines from Indian indicines

```{r}
read_structure(dataset_thinning = "20cattle.onepercent", k = 3) %>% 
  filter(variable == "X1") %>%
  filter(value != 0) %>% 
  arrange(desc(value), continent, region, population) %>% 
  select(variable, value, everything())


```


### K = 4

```{r}

read_structure(dataset_thinning = "20cattle.onepercent", k = 4) %>%
  mutate(
    color_var =
      case_when(
        # Asian taurines
        variable == "X1" ~ "#5f1f29",
        # African taurines
        variable == "X2" ~ "#fdaba3",
        # European taurine
        variable == "X3" ~ "#126180",
        # Indian indicine
        variable == "X4" ~ "#ff7436"
      )
  ) %>%
  ggstructure(wrap_var = "region", custom_color = TRUE) +
  labs(title = "Cattle & ancient, 5 samples per population, 1% SNPs: k = 4")

ggsave(here::here("figures/faststructure/20cattle.onepercent.K4.png"), width = 10, height = 7)
```

* Pink component separates out African taurines AND maroon component separates out Asian taurines

### K = 5

```{r}

read_structure(dataset_thinning = "20cattle.onepercent", k = 5) %>%
  mutate(
    color_var =
      case_when(
        # European taurine
        variable == "X1" ~ "#126180",
        # Indian indicine
        variable == "X2" ~ "#ff7436",
        # Asian indicine
        variable == "X3" ~ "#629d62",
        variable == "X4" ~ "#84d6d3",
        # African taurines
        variable == "X5" ~ "#fdaba3"
      )
  ) %>%
  ggstructure(wrap_var = "region", custom_color = TRUE) +
  labs(title = "Cattle & ancient, 5 samples per population, 1% SNPs: k = 5")

ggsave(here::here("figures/faststructure/20cattle.onepercent.K5.png"), width = 10, height = 7)
```

* Last component probably associated with technical error: only two (very different) individuals with > 0 % assignment

```{r, echo=TRUE}
read_structure(dataset_thinning = "20cattle.onepercent", k = 5) %>%
  arrange(variable, desc(value)) %>% 
  filter(variable == "X4") %>% 
  filter(value > 0) %>% 
  select(variable, value, everything())
```

* East Asian indicine in green (I think this is pretty noisy though - some Brahmans, which derive from Indian incine breeds, get assigned completely to the East Asian indicine cluster)
* Non-European taurines in pink

### K = 6

```{r}

read_structure(dataset_thinning = "20cattle.onepercent", k = 6) %>%
  mutate(
    color_var =
      case_when(
        # Indian indicine
        variable == "X1" ~ "#ff7436",
        # European taurine
        variable == "X3" ~ "#126180",
        # African taurines
        variable == "X4" ~ "#fdaba3",
        # Asian taurines
        variable == "X5" ~ "#5f1f29",        
        # Asian indicine
        variable == "X6" ~ "#629d62",
      )
  ) %>%
  ggstructure(wrap_var = "region", custom_color = TRUE) +
  labs(title = "Cattle & ancient, 5 samples per population, 1% SNPs: k = 6")

ggsave(here::here("figures/faststructure/20cattle.onepercent.K6.png"), width = 10, height = 7)
```

Overall, I K = 6 is pretty noisy. Several of the clusters don't make a ton of sense based on what we know a priori and/or completely separate out individuals within a breed

* Again, African taurines form their own cluster (pink)

```{r, echo=TRUE}
read_structure(dataset_thinning = "20cattle.onepercent", k = 6) %>%
  arrange(variable, desc(value)) %>% 
  filter(variable == "X4") %>% 
  filter(value > 0) %>% 
  select(variable, value, everything())
```

* Some Asian taurines and indicines in green 

```{r, echo=TRUE}
read_structure(dataset_thinning = "20cattle.onepercent", k = 6) %>%
  arrange(variable, desc(value)) %>% 
  filter(variable == "X6") %>% 
  filter(value > 0) %>% 
  select(variable, value, everything())
```

* Maroon cluster possibly noise - pulls out a few Mishima (a Korean taurine breed) and a few other Asian taurines/indicines

```{r, echo=TRUE}
read_structure(dataset_thinning = "20cattle.onepercent", k = 6) %>%
  arrange(variable, desc(value)) %>% 
  filter(variable == "X5") %>% 
  filter(value > 0) %>% 
  select(variable, value, everything()) %>% 
  left_join(sample_metadata %>% 
              select(international_id, avg_coverage, owner))
```

* Last component almost certainly associated with technical error: only one individual with > 0 % assignment

```{r, echo=TRUE}
read_structure(dataset_thinning = "20cattle.onepercent", k = 6) %>%
  arrange(variable, desc(value)) %>% 
  filter(variable == "X2") %>% 
  filter(value > 0) %>% 
  select(variable, value, everything())
```

## `5cattle.onepercent` {.tabset}

* Model complexity that maximizes marginal likelihood = 2
* Model components used to explain structure in data = 3

### K = 2

```{r}

read_structure(dataset_thinning = "5cattle.onepercent", k = 2) %>% 
  mutate(color_var =
           case_when(
             # European taurine
             variable == "X2" ~ "#126180",
             # Indian indicine
             variable == "X1" ~ "#ff7436"
           )) %>% 
  ggstructure(wrap_var = "region", custom_color = TRUE) +
  labs(title = "Cattle & ancient, 5 samples per population, 1% SNPs: k = 2")

ggsave(here::here("figures/faststructure/5cattle.onepercent.K2.png"), width = 10, height = 7)
```

### K = 3

* Components appear to be taurines (dark blue) and indicines (orage)
    + Other component associated with Siberian auroch (light blue)

```{r}

read_structure(dataset_thinning = "5cattle.onepercent", k = 3) %>%
  mutate(color_var =
           case_when(
             # European taurine
             variable == "X2" ~ "#126180",
             # Indian indicine
             variable == "X3" ~ "#ff7436",
             # auroch
             variable == "X1" ~ "#84d6d3")) %>%
  ggstructure(wrap_var = "region", custom_color = TRUE) +
  labs(title = "Cattle & ancient, 5 samples per population, 1% SNPs: k = 3")

ggsave(here::here("figures/faststructure/5cattle.onepercent.K3.png"), width = 10, height = 7)
```

### K = 4

* First three components appear to be taurines (dark blue), Indian indicines (orange), and East Asian indicines (green)
    + Barely visible, but the 4th component separates out Derbyshire auroch (likely related to technical error; light blue)

```{r}

read_structure(dataset_thinning = "5cattle.onepercent", k = 4) %>%
  mutate(
    color_var =
      case_when(
        # European taurine
        variable == "X1" ~ "#126180",
        # Indian indicine
        variable == "X2" ~ "#ff7436",
        # Asian indicine
        variable == "X3" ~ "#629d62",
        # auroch
        variable == "X4" ~ "#84d6d3"
      )
  ) %>%
  ggstructure(wrap_var = "region", custom_color = TRUE) +
  labs(title = "Cattle & ancient, 5 samples per population, 1% SNPs: k = 4")

ggsave(here::here("figures/faststructure/5cattle.onepercent.K4.png"), width = 10, height = 7)
```

# Outgroup datasets

## `allyak.onepercent` {.tabset}

* Model complexity that maximizes marginal likelihood = 2
* Model components used to explain structure in data = 3

### K = 2

```{r}

read_structure(dataset_thinning = "allyak.onepercent", k = 2) %>% 
  mutate(
    color_var =
      case_when(
        # Weird Jinchuan component
        variable == "X1" ~ "#5f1f29",
        # most yaks
        variable == "X2" ~ "#c6b7d5"
      )
  ) %>%
  ggstructure(wrap_var = "population", custom_color = TRUE) +
  labs(title = "All yaks, 1% SNPs: k = 2")

ggsave(here::here("figures/faststructure/allyak.onepercent.K2.png"), width = 10, height = 7)
```

* Second cluster in Jinchuan yaks probably associated with technical error - all of them come from the same university

```{r}
read_structure(dataset_thinning = "allyak.onepercent", k = 2) %>% 
  filter(variable == "X1") %>% 
  filter(value > 0.05) %>% 
  arrange(desc(value)) %>% 
  left_join(sample_metadata %>% 
              select(international_id, owner, avg_coverage))
```

### K = 3

```{r}

read_structure(dataset_thinning = "allyak.onepercent", k = 3) %>% 
  mutate(
    color_var =
      case_when(
        # Weird Jinchuan component
        variable == "X1" ~ "#5f1f29",
        # most yaks
        variable == "X2" ~ "#c6b7d5",
        variable == "X3" ~ "#e74a2f"       
      )
  ) %>%
  ggstructure(wrap_var = "population", custom_color = TRUE) +
  labs(title = "All yaks, 1% SNPs: k = 3")

ggsave(here::here("figures/faststructure/allyak.onepercent.K3.png"), width = 10, height = 7)
```

* Again, other 2 clusters likely technical error

```{r}
read_structure(dataset_thinning = "allyak.onepercent", k = 3) %>% 
  filter(variable %in% c("X1", "X3")) %>% 
  filter(value > 0.05) %>% 
  arrange(desc(value)) %>% 
  left_join(sample_metadata %>% 
              select(international_id, owner, avg_coverage))
```

## `allgaurgayal.onepercent` {.tabset}

* Model complexity that maximizes marginal likelihood = 2
* Model components used to explain structure in data = 1

### K = 2

```{r}
read_structure(dataset_thinning = "allgaurgayal.onepercent", k = 2) %>%
  mutate(
    color_var =
      case_when(
        # Banteng, gaur, gayal
        variable == "X1" ~ "#fedb11",
        # Yaks
        variable == "X2" ~ "#5da19a"
      )
  ) %>%
  ggstructure(wrap_var = "population", custom_color = TRUE) +
  labs(title = "All gaur and gayal, 1% SNPs: k = 2")

ggsave(here::here("figures/faststructure/allgaurgayal.onepercent.K2.png"), width = 10, height = 7)
```


# Commentary

* Within cattle datasets, I think using 5 samples per population rather than 20 gives results that are more interpretable/less variable & noisy
* Within taurines
    + Non-European (African, Asian) taurines make up their own cluster in `20all.onepercent` K = 6 and `20cattle.onepercent` K = 5
* Within indicines
  + East Asian indicines often form their own component
* Something weird with Jinchuan yaks
* No evidence for structure within banteng