---
title: "Phasing"
author: "Harly Durbin"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: show
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(tidylog)
library(magrittr)
```

# Notes & questions

* Sample sex determination?
    + ~~PLINK sex check~~
    + `bcftools guess_ploidy`: need coordinates of PAR 
        - ["The size of the PAR on the ARS-UCD1.2 reference genome assembly was 5,708,626 bp"](https://doi.org/10.3168/jds.2018-15638)
        - PAB at ChrX:133,300,518
    + Coverage on the sex chromosomes
        - True females may show coverage on Y due to mis-mapping of PAR?
    + Heterozygosity on the X
    
    
plink --vcf data/derived_data/joint_genotyping/remove_samples/remove_samples.X.vcf.gz --make-bed --double-id --cow --chr X -set-missing-var-ids @:#\$1\$2 --threads 12 --from-bp 1 --to-bp 133300518 --out test_x

plink --bfile test_x --double-id --cow --threads 12 --impute-sex --make-bed --out impute_sex
    
# Setup

```{r}
sample_metadata <- read_csv(here::here("data/derived_data/metadata/bovine_demo.sample_metadata.csv"))
```

```{r}
sex_impute <-
  read_rds(here::here("data/derived_data/metadata/coverage/cov_detail.rds")) %>% 
  filter(chr %in% c("X", "Y")) %>% 
  select(lab_id:mean, -total, -sid) %>% 
  tidyr::pivot_wider(values_from = mean, names_from = chr) %>% 
  right_join(sample_metadata %>% 
               select(lab_id:sex, bio_sample, species, population, avg_coverage)) %>% 
  select(lab_id, international_id, bio_sample, sex, population, species, everything()) %>% 
  # I know already that my ancient samples are both males
  mutate(sex = if_else(population == "ancient", "M", sex))
  
  
```

# Sex determination

## How many samples with no sex listed?

```{r}
sex_impute %>% 
  filter(is.na(sex)) %>% 
  group_by(population) %>% 
  tally(sort = TRUE)
```

## Based on coverage of sex chromosomes

```{r}
sex_impute %>%
  tidyr::pivot_longer(cols = c("X", "Y"), names_to = "chr") %>%
  mutate(sex = if_else(is.na(sex), "U", sex)) %>%
  ggplot(aes(x = value, y = avg_coverage, color = sex)) +
  geom_point(alpha = 0.4) +
  scale_color_manual(values = c("F" = "#F8766D",
                                "M" = "#00BFC4",
                                "U" = "lightgrey")) +
  theme_bw() +
  labs(x = "Average coverage on the sex chromosome", y = "Average coverage across all autosomes", color = "Sex") +
  facet_wrap(~ chr)

#ggsave(here::here("figures/phasing/xy_cov.all.png"))
```

```{r, fig.height=12, fig.width=6}
sex_impute %>%
  tidyr::pivot_longer(cols = c("X", "Y"), names_to = "chr") %>%
  mutate(sex = if_else(is.na(sex), "U", sex),
         species_group = 
           case_when(
             species %in% c("bos primigenius", "bos taurus", "composite", "bos indicus") ~ "Cattle",
             species %in% c("bos grunniens", "bos mutus") ~ "Yaks",
             TRUE ~ stringr::str_to_sentence(species)
           )) %>%
  ggplot(aes(x = value, y = avg_coverage, color = sex)) +
  geom_point(alpha = 0.4) +
  scale_color_manual(values = c("F" = "#F8766D",
                                "M" = "#00BFC4",
                                "U" = "lightgrey")) +
  theme_bw() +
  labs(x = "Average coverage on the sex chromosome", y = "Average coverage across all autosomes", color = "Sex") +
  facet_grid(species_group ~ chr)

#ggsave(here::here("figures/phasing/xy_cov.all.png"))
```

### Correlation between X coverage, Y coverage, and autosomal coverage

#### Known females

```{r}
sex_impute %>% 
  filter(sex == "F") %>% 
  select(X, Y, avg_coverage) %>% 
  corrr::correlate(diagonal = 1)
```
#### Known males

```{r}
sex_impute %>% 
  filter(sex == "M") %>% 
  select(X, Y, avg_coverage) %>% 
  corrr::correlate(diagonal = 1)
```

### Assign likely sex based on (mean coverage on X or Y/mean coverage on autosomes)

```{r}
sex_impute %>%
  filter(!is.na(sex)) %>%
  mutate(y_vs_auto = Y / avg_coverage,
         x_vs_auto = X / avg_coverage) %>%
  group_by(sex) %>%
  summarise(
    `Mean (Y/autosomes)` = mean(y_vs_auto),
    `Mean (X/autosomes)` = mean(x_vs_auto),
    `SD (Y/autosomes)` = sd(y_vs_auto),
    `SD (X/autosomes)` = sd(x_vs_auto)
  ) %>%
  select(Sex = sex, contains("X/"), contains("Y/"))

```

```{r}

sex_impute %<>% 
  mutate(sex = if_else(is.na(sex), "U", sex),
        y_vs_auto = Y / avg_coverage,
        x_vs_auto = X / avg_coverage,
        sex_cov =
          case_when(x_vs_auto >= (0.9172331 - 0.12504436) ~ "F",
                    TRUE ~ "M"))
```

## Based on `bcftools +guess-ploidy`

```{r}
sex_impute %<>%
  left_join(purrr::map(.x = c("GT", "PL"), ~ read_table2(
    here::here(
      glue::glue(
        "data/derived_data/joint_genotyping/impute_sex/bovine_demo.guess_ploidy.{.x}.txt"
      )
    ),
    col_names = c("international_id", glue::glue("sex_{.x}"))
  )) %>%
    reduce(full_join))



```

## Based on X chromosome heterozygosity using `plink --impute-sex`

```{r}
sex_impute %<>%
  left_join(
    read_table2(
      here::here(
        "data/derived_data/joint_genotyping/impute_sex/impute_sex.sexcheck"
      )
    ) %>%
      janitor::clean_names() %>%
      select(international_id = fid, sex_het = snpsex, f) %>%
      mutate(sex_het =
               case_when(
                 sex_het == 0 ~ "U",
                 sex_het == 1 ~ "M",
                 sex_het == 2 ~ "F"
               ))
  )
```


## Final determination based on all available information

```{r}
sex_impute %<>% 
  select(international_id, contains("sex"), x_vs_auto, y_vs_auto, f, everything())
```


* Quite a few `GT` inferred sexes disagree with other lines of evidence

```{r, fig.height=12, fig.width=6}
sex_impute %>%
  arrange(sex, sex_cov, sex_GT, sex_PL, sex_het) %>%
  mutate(international_id = forcats::fct_inorder(international_id)) %>% 
  select(international_id, contains("sex")) %>%
  tidyr::pivot_longer(cols = contains("sex"), names_to = "metric") %>%
  ggplot(aes(x = metric, y = international_id, fill = value)) +
  geom_tile() +
  scale_fill_manual(values = c("F" = "#F8766D",
                                "M" = "#00BFC4",
                                "U" = "lightgrey")) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())

ggsave(here::here("figures/phasing/sex_consensus.png"))
```

---

```{r}
sex_impute %>% 
  filter(x_vs_auto > 0.8) %>% 
  filter(sex == "M") %>% 
  pull(international_id) %>% 
  datapasta::vector_construct()
```

```{r}
sex_impute %>% 
  filter(0.55 > x_vs_auto) %>% 
  filter(sex == "F") 
```

```{r}
sex_impute %<>% 
  mutate(imputed_sex = 
           case_when(
             international_id %in% c("SIMDEUF000938065319", "UMCUSAM000000020087", "UMCUSAM000000033784", "UMCUSAM000000196770", "UMCUSAM000000198523", "UMCUSAM000000202633", "UMCUSAM000000204540", "UMCUSAM000000204542", "UMCUSAU000000194531", "UMCUSAU000000194532", "UMCUSAU000000194533", "UMCUSAU000000194535", "UMCUSAU000000194704") ~ "F",
             international_id %in% c("UMCUSAU000000205014") ~ "M",
             sex %in% c("M", "F") ~ sex,
             TRUE ~ sex_cov
           ))
```

```{r}
sex_impute %>% 
  group_by(imputed_sex) %>% 
  tally()
```

### Output guide for PLINK 

```{r}
sex_impute %>% 
  mutate(fid = international_id) %>% 
  select(international_id, fid, imputed_sex) %>% 
  write_tsv(here::here("data/derived_data/joint_genotyping/impute_sex/imputed_sexes.txt"), col_names = FALSE)
```


