---
title: "PCA"
author: "Harly Durbin"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(magrittr)
library(purrr)
library(dplyr)
library(stringr)
library(ggplot2)

source(here::here("source_functions/plot_pca.R"))
```

# Notes & questions

* [Eigensoft population lists vs. projection](https://compvar-workshop.readthedocs.io/en/latest/contents/02_pca/pca.html#population-lists-vs-projection)
* Maybe need to specify shapes manually

---

Possible metadata edits:

* Weird Jinchuan yaks
* Sichuan indigenous is admixed

## Dataset key

* Which samples

```{r}
tibble::tribble(
  ~ Name, ~ Description,
  "full", "All species. All available samples.",
  "20all", "All species. Up to 20 (ranked by coverage) samples per population label.",
  "20cattle", "Bos taurus, B.indicus, B. primigenius, composite. Up to 20 (ranked by coverage) samples per population label.",
  "5cattle", "Bos taurus, B.indicus, B. primigenius, composite. Up to 5 (ranked by coverage) samples per population label.",
  "5taurine", "Bos taurus, B. primigenius. Up to 5 (ranked by coverage) samples per population label.",
  "allyak", "B. grunniens and B. mutus. All available samples.",
  "allgayalgaur", "B. frontalis and B. gaurus. All available samples.",
  "allbanteng", "B. javanicus. All available samples."
) %>% 
  kableExtra::kable("html") %>% 
  kableExtra::kable_styling(full_width = F)
```

* Thinning parameter

```{r}
tibble::tribble(
  ~ Name, ~ Description,
  "superthin", "Randomly retain 0.001% of SNPs from each chromosome",
  "onepercent", "Randomly retain 0.01% of SNPs from each chromosome"
) %>%
  kableExtra::kable("html") %>% 
  kableExtra::kable_styling(full_width = F)
```


# Setup

```{r}
sample_metadata <- 
  read_csv(here::here("data/derived_data/metadata/bovine_demo.sample_metadata.csv")) %>% 
  mutate(international_id = if_else(international_id == "CPC98_Bprimigenius_EnglandDerbyshire_5936", "ancient_derbyshire", international_id))
```

# Full datasets {.tabset}

## `full`

```{r, fig.width = 7, fig.height=5}
plot_pca(dataset_thinning = "projectfull.onepercent", pc1 = 1, pc2 = 2, color_var = species, shape_var = continent, title_string = "All species, all samples, 1% SNPs")
```

```{r, fig.width = 7, fig.height=5}
ggscree(dataset_thinning = "projectfull.onepercent", title_string = "All species, all samples, 1% SNPs")
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
purrr::map2(
  .x = seq(1,9, by = 2),
  .y = seq(2,10, by = 2),
  ~ plot_pca(
    dataset_thinning = "projectfull.onepercent",
    pc1 = .x, pc2 = .y,
    color_var = species,
    shape_var = continent,
    title_string = "All species, all samples, 1% SNPs") %>% 
    ggplot2::ggsave(
      filename = here::here(glue::glue("figures/smartpca/projectfull.onepercent.PC{.x}_PC{.y}.png")),
      width = 7,
      height = 5
    ))
```


## `20all`

```{r, fig.width = 7, fig.height=5}
plot_pca(dataset_thinning = "project20all.onepercent", pc1 = 1, pc2 = 2, color_var = species, shape_var = continent, title_string = "All species, 20 samples per population, 1% SNPs")
```

```{r, fig.width = 7, fig.height=5}
ggscree(dataset_thinning = "project20all.onepercent", title_string = "All species, 20 samples per population, 1% SNPs")
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
purrr::map2(
  .x = seq(1,9, by = 2),
  .y = seq(2,10, by = 2),
  ~ plot_pca(
    dataset_thinning = "project20all.onepercent",
    pc1 = .x, pc2 = .y,
    color_var = species,
    shape_var = continent,
    title_string = "All species, 20 samples per population, 1% SNPs") %>% 
    ggplot2::ggsave(
      filename = here::here(glue::glue("figures/smartpca/project20all.onepercent.PC{.x}_PC{.y}.png")),
      width = 7,
      height = 5
    ))
```

## `5regionall`

```{r, fig.width = 7, fig.height=5}
plot_pca(dataset_thinning = "project5regionall.onepercent", pc1 = 1, pc2 = 2, color_var = species, shape_var = continent, title_string = "All species, 5 samples per region per population, 1% SNPs")
```

```{r, fig.width = 7, fig.height=5}
ggscree(dataset_thinning = "project5regionall.onepercent", title_string = "All species, 20 samples per population, 1% SNPs")
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
purrr::map2(
  .x = seq(1,9, by = 2),
  .y = seq(2,10, by = 2),
  ~ plot_pca(
    dataset_thinning = "project5regionall.onepercent",
    pc1 = .x, pc2 = .y,
    color_var = species,
    shape_var = continent,
    title_string = "All species, 5 samples per region per population, 1% SNPs") %>% 
    ggplot2::ggsave(
      filename = here::here(glue::glue("figures/smartpca/project5regionall.onepercent.PC{.x}_PC{.y}.png")),
      width = 7,
      height = 5
    ))
```

# Cattle datasets {.tabset}

## `20cattle`

```{r, fig.width = 7, fig.height=5}
plot_pca(dataset_thinning = "project20cattle.onepercent", pc1 = 1, pc2 = 2, color_var = species, shape_var = continent, title_string = "Cattle & ancient, 20 samples per population, 1% SNPs")
```

```{r, fig.width = 7, fig.height=5}
ggscree(dataset_thinning = "project20cattle.onepercent", title_string = "Cattle & ancient, 20 samples per population, 1% SNPs")
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
purrr::map2(
  .x = seq(1,9, by = 2),
  .y = seq(2,10, by = 2),
  ~ plot_pca(
    dataset_thinning = "project20cattle.onepercent",
    pc1 = .x, pc2 = .y,
    color_var = species,
    shape_var = continent,
    title_string = "Cattle & ancient, 20 samples per population, 1% SNPs") %>% 
    ggplot2::ggsave(
      filename = here::here(glue::glue("figures/smartpca/project20cattle.onepercent.PC{.x}_PC{.y}.png")),
      width = 7,
      height = 5
    ))
```

## `5cattle`

* Some Asian taurines mislabeled as composites and vice versa?

```{r, fig.width = 7, fig.height=5}
plot_pca(dataset_thinning = "project5cattle.onepercent", pc1 = 1, pc2 = 2, color_var = species, shape_var = continent, title_string = "Cattle & ancient, 5 samples per population, 1% SNPs")
```

```{r, fig.width = 7, fig.height=5}
ggscree(dataset_thinning = "project5cattle.onepercent", title_string = "Cattle & ancient, 5 samples per population, 1% SNPs")
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
purrr::map2(
  .x = seq(1,9, by = 2),
  .y = seq(2,10, by = 2),
  ~ plot_pca(
    dataset_thinning = "project5cattle.onepercent",
    pc1 = .x, pc2 = .y,
    color_var = species,
    shape_var = continent,
    title_string = "Cattle & ancient, 5 samples per population, 1% SNPs") %>% 
    ggplot2::ggsave(
      filename = here::here(glue::glue("figures/smartpca/project5cattle.onepercent.PC{.x}_PC{.y}.png")),
      width = 7,
      height = 5
    ))
```

### Both ancient samples cluster with taurine breeds native to the Asian steppe

```{r, fig.width = 7, fig.height=8}
read_vec("project5cattle.onepercent") %>%
  mutate(
    color_var =
      case_when(
        population %in% c("yakutian", "mongolian", "chaidamu yellow", "kazakh", "tibetan") ~ stringr::str_to_title(population),
        international_id == "A2494_Yenisei_River" ~ "Ancient (Siberian)",
        international_id == "ancient_derbyshire" ~ "Ancient (Derbyshire)",
        TRUE ~ "All others"),
    color_var = forcats::fct_relevel(as.factor(color_var), "Ancient (Siberian)", "Ancient (Derbyshire)", "Yakutian", "Mongolian", "Chaidamu Yellow", "Kazakh", "Tibetan", "All others"),
    alpha_var = if_else(color_var == "All others", 0.2, 0.6),
    size_var = if_else(population == "ancient", 4, 3),
    species = stringr::str_to_sentence(species)
  ) %>%
  ggplot(aes(x = PC1,
             y = PC2,
             color = color_var,
             shape = species,
             alpha = alpha_var,
             size = size_var)) +
  geom_point() +
  scale_alpha_identity() +
  scale_size_identity() +
  scale_color_manual(
    values = c(
      "Ancient (Siberian)" = "#126180",
      "Ancient (Derbyshire)" = "#629d62",
      "Yakutian" = "#84d6d3",
      "Mongolian" = "#ee5863",
      "Chaidamu Yellow" = "#fedb11",
      "Kazakh" = "#ff743d",
      "Tibetan" = "#5f1f29",
      "All others" = "lightgrey"
      )
  ) +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  theme_classic() +
  labs(shape = "Species",
       color = "Population",
       title = "Cattle & ancient, 5 samples per population, 1% SNPs",
       x = "PC 1: 51.1%",
       y = "PC 2: 14.2%") +
  ggforce::facet_zoom(xlim = c(-.045, -0.01),
             ylim = c(-.015, .025),
             zoom.size = 0.5,
             split = FALSE,
             horizontal = FALSE)

ggsave(here::here("figures/smartpca/project5cattle_asiansteppe_cutout.onepercent.PC1_PC2.png"), width = 7, height = 5)
```

## `5taurine`

* Sichuan indigenous probably not a taurine cow or is admixed

```{r, fig.width = 7, fig.height=5}
plot_pca(dataset_thinning = "project5taurine.onepercent", pc1 = 1, pc2 = 2, color_var = region, shape_var = continent, title_string = "Taurine cattle & ancient, 5 samples per population, 1% SNPs")
```

```{r, fig.width = 7, fig.height=5}
ggscree(dataset_thinning = "project5taurine.onepercent", title_string = "Taurine cattle & ancient, 5 samples per population, 1% SNPs")
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
purrr::map2(
  .x = seq(1,9, by = 2),
  .y = seq(2,10, by = 2),
  ~ plot_pca(
    dataset_thinning = "project5taurine.onepercent",
    pc1 = .x, pc2 = .y,
    color_var = region,
    shape_var = continent,
    title_string = "Taurine cattle & ancient, 5 samples per population, 1% SNPs") %>% 
    ggplot2::ggsave(
      filename = here::here(glue::glue("figures/smartpca/project5taurine.onepercent.PC{.x}_PC{.y}.png")),
      width = 7,
      height = 5
    ))
```

# Outgroup datasets {.tabset}

## `allbanteng`

```{r, fig.width = 7, fig.height=5}
plot_pca(dataset_thinning = "allbanteng.onepercent", pc1 = 1, pc2 = 2, color_var = international_id, title_string = "All banteng, 1% SNPs")
```

```{r, fig.width = 7, fig.height=5}
ggscree(dataset_thinning = "allbanteng.onepercent", title_string = "All banteng, 1% SNPs")
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
purrr::map2(
  .x = seq(1,9, by = 2),
  .y = seq(2,10, by = 2),
  ~ plot_pca(
    dataset_thinning = "allbanteng.onepercent",
    pc1 = .x, pc2 = .y,
    color_var = international_id,
    title_string = "All banteng, 1% SNPs") %>% 
    ggplot2::ggsave(
      filename = here::here(glue::glue("figures/smartpca/allbanteng.onepercent.PC{.x}_PC{.y}.png")),
      width = 7,
      height = 5
    ))
```

## `allgaurgayal`

```{r, fig.width = 7, fig.height=5}
plot_pca(dataset_thinning = "allgaurgayal.onepercent", pc1 = 1, pc2 = 2, color_var = population, title_string = "All gaur & gayal, 1% SNPs")
```

```{r, fig.width = 7, fig.height=5}
ggscree(dataset_thinning = "allgaurgayal.onepercent", title_string = "All gaur & gayal, 1% SNPs")
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
purrr::map2(
  .x = seq(1,9, by = 2),
  .y = seq(2,10, by = 2),
  ~ plot_pca(
    dataset_thinning = "allgaurgayal.onepercent",
    pc1 = .x, pc2 = .y,
    color_var = population,
    title_string = "All gaur & gayal, 1% SNPs") %>% 
    ggplot2::ggsave(
      filename = here::here(glue::glue("figures/smartpca/allgaurgayal.onepercent.PC{.x}_PC{.y}.png")),
      width = 7,
      height = 5
    ))
```

## `allyak`

* Probably 3 mislabeled "yak jinchuan"

```{r, fig.width = 7, fig.height=5}
plot_pca(dataset_thinning = "allyak.onepercent", pc1 = 1, pc2 = 2, color_var = population, title_string = "All domestic & wild yak, 1% SNPs")
```

```{r, fig.width = 7, fig.height=5}
ggscree(dataset_thinning = "allyak.onepercent", title_string = "All domestic & wild yak, 1% SNPs")
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
purrr::map2(
  .x = seq(1,9, by = 2),
  .y = seq(2,10, by = 2),
  ~ plot_pca(
    dataset_thinning = "allyak.onepercent",
    pc1 = .x, pc2 = .y,
    color_var = population,
    title_string = "All domestic & wild yak, 1% SNPs") %>% 
    ggplot2::ggsave(
      filename = here::here(glue::glue("figures/smartpca/allyak.onepercent.PC{.x}_PC{.y}.png")),
      width = 7,
      height = 5
    ))
```

# Commentary

* Previously, the Siberian auroch defined PC2 and the Derbyshire auroch clustered closely with European cattle. Now that we have many more Asian cattle represented, the two aurochs are much closer together and are clustered within breeds found on the Asian steppe.
* In cattle datasets, PC1 is taurine vs. indicine and PC2 appears to be Indian indicine vs. East Asian indicine (though quite a lot of variation within East Asian indicine and composite)
* No idea what the deal with weird 3 Jinchuan yaks, probably need to re-run without them