---
title: "Treemix"
author: "Harly Durbin"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(stringr)
library(magrittr)
library(rlang)
library(purrr)
library(ggplot2)
library(tidylog)

source(here::here("source_functions/popcorn_treemix.R"))
```

# Notes & questions

* ["Because we only use a single individual per taxon, we need to use the flag -noss. Otherwise, treemix will apply a correction for small sample size which is too conservative."](https://speciationgenomics.github.io/Treemix/)
*[ipyrad-analysis toolkit: treemix](https://ipyrad.readthedocs.io/en/latest/API-analysis/cookbook-treemix.html)

# Setup

```{r}
sample_metadata <- 
  read_csv(here::here("data/derived_data/metadata/bovine_demo.sample_metadata.csv")) %>% 
  mutate(international_id = if_else(international_id == "CPC98_Bprimigenius_EnglandDerbyshire_5936", "ancient_derbyshire", international_id))
```

```{r}
theme_hj_treemix <- function(gg) {
  gg <-
    gg +
    ggplot2::theme_classic() +
    ggplot2::theme(
      axis.text = ggplot2::element_text(size = 22),
      axis.title.x = ggplot2::element_text(size = 26,
                                           margin = margin(
                                             t = 15,
                                             r = 0,
                                             b = 0,
                                             l = 0
                                           )),
      axis.line.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      axis.text.y = ggplot2::element_blank(),
      axis.title.y = ggplot2::element_blank(),
      legend.position = c(0.2, 0.4),
      legend.title = element_text(size = 24),
      legend.text = element_text(size = 20)
    ) +
    guides(color = guide_legend(title = "Migration weight",
                                override.aes = list(size = 10))) +
    labs(x = "Drift parameter")
  
  return(gg)
}
```



# Modified full results {.tabset}

* Removed several individuals/populations
    + Recent American/Australian composite breeds (Santa Gertrudis, Beefmaster, and Droughtmaster)
    + "East African Zebu" with quality issues in fastSTRUCTURE and PCA
* Consolidated several populations 
    + All domestic yak breeds
    + All gayal populations
    + All Podolian Steppe breeds (Podolica, Maremmana, Boskarin, Sikias)
    + Combined Lowline Angus with Angus
    + Combined miniature Hereford with Hereford
* Removed all remaining populations with fewer than 5 individuals **excluding** Asian steppe breeds (Chaidamu Yellow) and the consolidated Podolian Steppe population 

The resulting dataset has 2,167 individuals:

```{r}
read_table2(here::here("data/derived_data/treemix/define_cluster.full.onepercent.txt"), col_names = c("id", "fid", "Population")) %>% 
  group_by(Population) %>% 
  tally(sort = TRUE)
```

```{r, eval = FALSE}
read_table2(
  here::here(
    "data/derived_data/treemix/define_cluster.full.onepercent.txt"
  ),
  col_names = c("international_id", "fid", "tmpop")
) %>%
  left_join(sample_metadata %>%
              select(international_id, species, continent, region, population)) %>%
  mutate(
    continent = str_to_title(continent),
    continent = forcats::fct_relevel(as.factor(continent),
                                     "europe",
                                     "africa",
                                     "asia",
                                     "americas"),
    region = case_when(
      region == "india and pakistan" ~ "India & Pakistan",
      region == "japan and korea" ~ "Japan & Korea",
      TRUE ~ str_to_title(region)
    ),
    region = forcats::fct_relevel(
      as.factor(region),
      "British Isles",
      "France",
      "Continental Europe",
      "Scandinavia",
      "Iberia",
      "Italy",
      "Podolian Steppe",
      "West Africa",
      "East Africa",
      "Middle East",
      "India & Pakistan",
      "Japan & Korea",
      "Southeast Asia",
      "Asian Steppe",
      "Northeast China",
      "East Central China",
      "South Central China",
      "Southeast China",
      "Americas"
    )
  ) %>% 
  arrange(region, tmpop) %>% 
  distinct(tmpop) %>% 
  write_tsv(here::here("data/derived_data/treemix/output/full.onepercent/poporder.full.onepercent.txt"), col_names = FALSE)
```

## Base tree

```{r, fig.width=10, fig.height=12}
base <- read_treemix(here::here("data/derived_data/treemix/output/full.onepercent/treemix.full.onepercent.base"))
```


```{r, fig.width=18, fig.height=18}
plot_treemix(base, text_size = 8) %>% 
  theme_hj_treemix() +
  expand_limits(x = 0.099) 

ggsave(here::here("figures/treemix/full.onepercent.base.png"), width = 18, height = 18)
```

## m = 1

```{r, fig.width=10, fig.height=12}
m1 <- read_treemix(here::here("data/derived_data/treemix/output/full.onepercent/treemix.full.onepercent.1"))
```


```{r, fig.width=18, fig.height=18}
plot_treemix(m1, text_size = 8) %>% 
  theme_hj_treemix() +
  expand_limits(x = 0.099)

ggsave(here::here("figures/treemix/full.onepercent.m1.png"), width = 18, height = 18)
```

# Commentary
